Type-Safe Modular Parsing

### 2017-06-07 SLE_17/sections/Inheritance.tex
12 pages.

\subsection{Extensible Parsing via Inheritance}

@@

To illustrate such extensibility, we continue with the old
example, and introduce variables as a new case.
It is easy to extend the corresponding OO AST in a modular way:

\lstinputlisting[linerange=7-9]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_SIMPLE_LAM

For code reuse, we want to define a new parser that extends
the old one. Here one may quickly come up
with the following attempt, where a new parser is defined for \inlinecode{Var}, then composed
with \inlinecode{pExpr}:

\lstinputlisting[linerange=13-16]{code/src/papercode/Sec3Inheritance/Code1.scala}% APPLY:linerange=INHERITANCE_BAD_ATTEMPT

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modifying existing code sacrifices separate compilation, and hence breaks semantic modularity.

\subsection{Overriding for Extensibility}\label{subsec:overriding}

It is actually quite simple to let \inlinecode{pExpr} cover the newly
extended case without modifying existing code. Method overriding is a
standard feature which often comes with inheritance, and it allows us
to redefine an inherited method, such as \inlinecode{pExpr}. We can
build the new parser which correctly parses \inlinecode{"1 + x"}
through overriding:

@@

\vspace{-.05in}
\paragraph{Independent Extensibility}
A nice feature of Scala is its support for the linearized-style multiple
inheritance on traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. Later we introduce language components
in Section~\ref{subsec:language-component} as a nice pattern.

In the above code, we use the \lstinline{super} keyword to refer to the old \inlinecode{pExpr}.
With multiple inheritance, Scala has a special syntax
\lstinline{super[T].x}, called \textit{static super reference}. It refers to the name
\inlinecode{x} in the specified parent trait \inlinecode{T}. With this we can reuse
the old parsers even if they have the same name.
\bruno{Independent extensibility cannot be explained like this. 
We need to see some code here! You need to expand the explanation. 
I suggest instead cutting on the ``wrong'' attempt in the beggining of this section, and instead 
have some brief discussion in the \emph{correct} code where we need to be careful.}

@@

\vspace{-.05in}
\paragraph{Conflicts and/or ambiguity} In a modular setting, conflicts and
ambiguity could be introduced to the grammar. In that case, the help parser combinators
can offer is quite restricted. Yet users can override those problematic methods to resolve
such conflicts, and rely on dynamic dispatch. We will discuss it in Section~\ref{subsec:overriding-rules}.

>>>>

@@

To illustrate such extensibility, we continue with the old
example, and introduce variables as a new case.
It is easy to extend the corresponding OO AST together with its parser in a modular way:

\begin{lstlisting}
class Var(x: String) extends Expr {
  def print = x
}
\end{lstlisting}
Here one may quickly define a new parser \lstinline|pVar| in \lstinline|AParser| for variables, and parse new expressions
with \lstinline{"pExpr ||| pVar"}.
Unfortunately, even \inlinecode{"1 + x"} cannot be parsed, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to \lstinline|pExpr| for parsing sub-expressions, whereas the newly added \inlinecode{pVar} is not observed, unless we replace all the occurrences of \lstinline|pExpr| with \lstinline{"pExpr ||| pVar"}.
Yet modifying existing code breaks semantic modularity.

\vspace{-.05in}
\paragraph{Overriding for Extensibility}
It is actually quite simple to let \inlinecode{pExpr} cover the newly
extended case without modifying existing code. Method overriding is a
standard feature which often comes with inheritance, and it allows us
to redefine an inherited method, such as \inlinecode{pExpr}. We can
build the new parser which correctly parses \inlinecode{"1 + x"}
through overriding:

@@

\vspace{-.05in}
\paragraph{Independent Extensibility}
A nice feature of Scala is its support for the linearized-style multiple
inheritance on traits~\cite{odersky2004overview}. This can be very helpful when
composing several languages, and to achieve independent
extensibility~\cite{odersky2005independently}. Suppose now we want to compose the parsers
for expressions from pre-defined languages \lstinline|LanguageA| and \lstinline|LanguageB| using alternative.
The new parser can be built by inheriting both parsers at the same time:
\begin{lstlisting}
trait LanguageA {...}
trait LanguageB {...}
trait LanguageC extends LanguageA with LanguageB {
  override def pExpr = super[LanguageA].pExpr ||| super[LanguageB].pExpr
}
\end{lstlisting}

The \lstinline{super[T].x} syntax in Scala, so called \textit{static super reference}, refers to the type or method
\inlinecode{x} in the parent trait \inlinecode{T}. Under multiple inheritance, it can be used to distinguish
the methods of the same name. Therefore in the new parser, we use \lstinline{super} to specify which
\lstinline{pExpr} we are referring to.

@@

\vspace{-.05in}
\paragraph{Conflicts and/or Ambiguity} In a modular setting, conflicts and
ambiguity could be introduced to the grammar. In that case, the help parser combinators
can offer is quite restricted. Yet users can override those problematic methods to resolve
such conflicts, and rely on dynamic dispatch. We will discuss it in Section~\ref{subsec:overriding-rules}.
